Pool Model

### 2025-08-21 pool_model.tex
update citation for cellular_raza
Restructure: relocate parts with mathematical formalism to the supplement section

\section{Introduction}
Motivation: general, flexible approach for bacterial systems that allows easily modify the system depending on processes on different scales.
Allows us to describe the processes within populations using different bacterial states as well as interactions between different populations using ODEs.\\
Existing models in Predictive Microbiology (Primary Models)~\cite{perez-rodriguez_predictive_2012}:
\begin{itemize}
    \item An overview of primary models used in food science can be found in~\cite{van_boekel_kinetic_2008}.
    \item Baranyi-Roberts Model~\cite{baranyi_modeling_1993, baranyi_dynamic_1994}
    \item modified Gompertz
    \item Buchanan et al.: three-phase linear model for lag phase, exponential growth phase, and stationary phase~\cite{buchanan_when_1997}
    \item McKellar two-compartment model with bacteria in growing and non-growing state~\cite{mckellar_heterogeneous_1997}
    \item Inactivation models
\end{itemize}
Secondary models:
\begin{itemize}
    \item Polynomial models 
    \item Square Root-Type Models, e.g., Ratkowski for temperature~\cite{ratkowsky_relationship_1982}
    \item The Cardinal Parameter Model~\cite{zwietering_decision_1992}
\end{itemize}
Comprehensive reviews about competition/cooperation bacterial mechanisms: \cite{ghoul_ecology_2016, stubbendieck_bacterial_2016, hibbing_bacterial_2010, west_social_2007}\\
\section{General Mathematical Formulation}
In the following section we demonstrate the application of this approach for different toy models describing the processes within and between bacteria communities.

@@

Motivated by this we divide a bacterial population into three pools (states) (~\ref{fig:SchematicRep}):
\begin{itemize}
    \item $x^1_1 (t) = L(t)$: the fraction of the population in the lag phase
    \item $x^2_1 (t) = G(t)$: the fraction of the population growing and dividing
    \item $x^3_1 (t) = D(t)$: the fraction of the population undergoing cell death
\end{itemize}

@@

To determine the function $\mbF(\mbx, \mbm, \mbp, \mbu)$ consider the flows between different states.

@@

or in matrix form it shows the linear dependency
    \end{pmatrix} \mbx .
\subsection{Resource Dependence}

@@

@@

        \label{eq:3pool_resource} 

@@

Rewriting the equations above, we get the following non-linear system:

@@

        The time dependence of the total bacterial count $N(t) = \sum_i \sum_k \mbx_i^k = L(t) + G(t) + D(t)$ is obtained as a sum of bacteria in all states.
        All curves were calculated using the model (\ref{eq:3pool_resource}) with parameters $\alpha=2.5$, $\mu'=0.5$ and $\beta=5$.

@@

\cmt{So now is it better to use flow to dormancy instead?}

@@

\subsubsection{Maxwell type of stress - strain relation}

@@

This system of equations defines the microenvironment $\mbm$ and bacteria state matrix $\mbx$ as
    \mbm = R, \hspace{1cm}
        L_A & L_B \\
        G_A & G_B 

@@

        The red curve describes the growth of the bacteria $A$ ($N_A = \sum_{k} x_A^k = L_A+G_A$) with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
        The blue curve describes the growth of the bacteria $B$ ($N_B = \sum_{k} x_B^k = L_B+G_B$) with model parameters $\lambda_B=5\cdot 10^{-2}$, $\alpha_B=2.5$.
        The purple line shows the total bacterial count of the system  $N(t)=\sum_i\sum_k\mbx_i^k=N_A+N_B$.

@@

In previous studies it was shown that \ac{glv} equations with one bacterial state can be sucessfully applied to predict the microbial community dynamics, determine different types of the pairwise interspecies interactions~\cite{dedrick_when_2023, stein_ecological_2013, venturelli_deciphering_2018, hoffmann_power_2007}.

@@

@@

        The two species system competing for the common nutritient $R$ with production of the toxin $T_B$ (brown line) calculated using Eq.~(\ref{eq:model_2sp_toxin}).

@@

        The red curve describes the growth of the bacteria $A$ ($N_A = \sum_{k} x_A^k = L_A+G_A$) with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
        The blue curve describes the growth of the bacteria $B$ ($N_B = \sum_{k} x_B^k = L_B+G_B$) with model parameters $\lambda_A=5\cdot 10^{-2}$, $\alpha_A=2.5$.

@@

We use \texttt{cellular\_raza}~\cite{Pleyer_cellular_raza_2023} to implement an \ac{abm}.

@@

@@

\subsection{Deriving the Pool Model}

@@

\subsection{Code Samples}

@@

\subsection{Full List of Parameters}

@@

\subsection{Homogenous and Inhomogenous Initial Cell Distributions}

>>>>

\newpage
%
%
%
%
% ================================================================================================
\section{Introduction}
Motivation: general, flexible approach for bacterial systems that allows easily modify the system depending on processes on different scales.
Allows us to describe the processes within populations using different bacterial states as well as interactions between different populations using ODEs.\\
%
%
Existing models in Predictive Microbiology (Primary Models)~\cite{perez-rodriguez_predictive_2012}:
\begin{itemize}
    \item An overview of primary models used in food science can be found in~\cite{van_boekel_kinetic_2008}.
    \item Baranyi-Roberts Model~\cite{baranyi_modeling_1993, baranyi_dynamic_1994}
    \item modified Gompertz
    \item Buchanan et al.: three-phase linear model for lag phase, exponential growth phase, and stationary phase~\cite{buchanan_when_1997}
    \item McKellar two-compartment model with bacteria in growing and non-growing state~\cite{mckellar_heterogeneous_1997}
    \item Inactivation models
\end{itemize}
%
Secondary models:
\begin{itemize}
    \item Polynomial models 
    \item Square Root-Type Models, e.g., Ratkowski for temperature~\cite{ratkowsky_relationship_1982}
    \item The Cardinal Parameter Model~\cite{zwietering_decision_1992}
\end{itemize}
%
%
Comprehensive reviews about competition/cooperation bacterial mechanisms: \cite{ghoul_ecology_2016, stubbendieck_bacterial_2016, hibbing_bacterial_2010, west_social_2007}\\
%
%

@@

Motivated by this we divide a bacterial population into three pools (states)~(\ref{fig:SchematicRep}) defining the fraction of the population in the lag phase $L(t)$, growing and dividing $G(t)$ and undergoing cell death $D(t)$.

@@

To determine the system of equations describing the system dynamics consider the flows between different states.

@@

\label{eq:ode_3pools}

@@

        \dot{G} = \lambda L + \alpha R G-\mu' G\\
        \dot{D} = \mu  L + \mu' G- \beta D\\
        \dot{R} = - \frac{\alpha}{N_t} R G
\label{eq:ode_3pool_resource} 

@@

@@

\label{eq:ode_3pool_resource2} 
\begin{figure}[h]

@@

        The time dependence of the total bacterial count $N(t) = L(t) + G(t) + D(t)$ is obtained as a sum of bacteria in all states.
        All curves were calculated using the model (\ref{eq:ode_3pool_resource2}) with parameters $\alpha=2.5$, $\mu'=0.5$ and $\beta=5$.

@@

@@

\paragraph{Maxwell type of stress - strain relation}

@@

@@

        The red curve describes the growth of the bacteria $A$ ($N_A = L_A+G_A$) with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
        The blue curve describes the growth of the bacteria $B$ ($N_B = L_B+G_B$) with model parameters $\lambda_B=5\cdot 10^{-2}$, $\alpha_B=2.5$.
        The purple line shows the total bacterial count of the system  $N(t)=N_A+N_B$.

@@

In previous studies it was shown that \ac{glv} equations with one bacterial state can be successfully applied to predict the microbial community dynamics, determine different types of the pairwise interspecies interactions~\cite{dedrick_when_2023, stein_ecological_2013, venturelli_deciphering_2018, hoffmann_power_2007}.

@@

For further comparison of the derived model with the matrix formulation of the \ac{glv} we refer the reader to the supplement materials~\ref{ssec:supplement1}.

@@

        The two species system competing for the common nutrient $R$ with production of the toxin $T_B$ (brown line) calculated using Eq.~(\ref{eq:model_2sp_toxin}).

@@

        The red curve describes the growth of the bacteria $A$ ($N_A = L_A+G_A$) with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
        The blue curve describes the growth of the bacteria $B$ ($N_B = L_B+G_B$) with model parameters $\lambda_A=5\cdot 10^{-2}$, $\alpha_A=2.5$.

@@

We use \texttt{cellular\_raza}~\cite{Pleyer2025} to implement an \ac{abm}.

@@

\section{Discussion}

@@

\subsection{General Mathematical Formalism}  \label{ssec:supplement1}
Using such matrix representation of the bacterial system, we can rewrite introduced above system in more compact form.
The three pool system described with equations~\ref{eq:ode_3pools} will be written as
    \dot \mbx  = A \mbx, \hspace{3mm} A = \begin{pmatrix}
\label{eq:x_repres_3pools}
With introduction of the limiting resource the system dynamics presented by the \acp{ode}~\ref{eq:ode_3pool_resource} will be rewritten in the following non-linear system:
\label{eq:x_repres_3pools_resource}
\label{eq:x_repres_3pools_resource2}
\paragraph{Generalized Lotka-Volterra}

This system of equations defines the microenvironment $\mbm$ and bacteria state matrix $\mbx$ as
\begin{equation}
    \mbm = R, \hspace{1cm}
    \mbx = \begin{pmatrix}
        L_A & L_B \\
        G_A & G_B 
    \end{pmatrix}.
\end{equation}
%

\subsection{ABM Details} \label{ssec:supplement2}

\subsubsection*{Deriving the Pool Model}

@@

\subsubsection*{Code Samples}

@@

\subsubsection*{Full List of Parameters}

@@

\subsubsection*{Homogenous and Inhomogenous Initial Cell Distributions}
